Verified Nanopasses for Compiling Conditionals

### 2025-08-11 paper.tex
add citation to continuation based PE
improvements from reviewers

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents an elegant family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that (1) no longer
contains Boolean expressions, (2) the condition expression of a
condition is a value, and (3) jumps are expressed using zero-arity
function calls. One of the key parts of that translation is in the
compilation of a conditional expression such as $(\IF{b}{e_1}{e_2})$,
where the expression $b$ is compiled using a special function
$\mathcal{B}$ that takes two more arguments, the generated code for
the two branches, that is, for the two continuations of $b$. Thus,
$\mathcal{B}$ can analyze the expression $b$ to decide how to generate
code for the $\mathsf{if}$, or whether an $\mathsf{if}$ is even
necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}

My first exposure to these ideas were in Kent Dybvig's compiler
course~\citep{Dybvig:2010aa} where he taught the techniques used in
the Chez Scheme compiler~\citep{Dybvig:2006aa}.  One of those
techniques is called destination-driven code
generation~\citep{Dybvig:1990aa}, which adds an extra parameter to the
compilation function to indicate how the context will use the result
value of an expression. The context may (1) ignore the result (and
only care about the expression's side effects), (2) branch based on
interpreting its result as a boolean, (3) assign the result to a
variable, or (4) output the result. These options are encoded as a
disjoint union.  Relating this to Olivier's transformation functions,
the $\mathcal{E}$ function is for contexts that need the actual value
and the $\mathcal{B}$ function is for contexts that branch on the
expression's result. Of course, a pair of functions is isomorphic to a
function on a disjoint union.
\[
(A → C) × (B → C) ≅ (A + B) → C
\]

The CompCert compiler \citep{Leroy:2006fe} (a compiler for C verified
in Coq) uses the one-function per context approach of Danvy in the
RTLgen pass that translates from the CminorSel intermediate language
to CompCert's register transfer language. The verified CakeML
compiler, on the other hand, employs a naive translation of Boolean
expressions and conditionals for the compilation of stackLang the
intermediate language into labLang~\citep{Kumar:2014aa}.

@@

This paper presents a proof of correctness in Agda for four nanopasses
that translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. The Agda code is available in the supplementary
material. The four compiler passes presented here correspond to the
following three passes in \emph{Essentials of Compilation}:
\begin{enumerate}
\item Remove Complex Operands
\item Explicate Control
\item Select Instructions
\end{enumerate}
The process of proving of correctness of the Explicate Control pass
made it clear to me that it has two distinct responsibilities that
deserve to be separated:
\begin{itemize}
\item Replacing \texttt{let} expressions with assignment statements.
\item Translating Boolean expressions and conditionals into
  \texttt{goto}-based control-flow.
\end{itemize}
(The RTLgen pass of CompCert also handles the above two
responsibilities.)  So in this paper, the Explicate Control pass is
split into two passes, the first named Lift Locals and the second that
keeps the name Explicate Control.
\begin{enumerate}
\item Remove Complex Operands
\item Lift Locals
\item Explicate Control
\item Select Instructions
\end{enumerate}
In compilers that do not take the nanopass approach, unrelated
responsibilities are sometimes combined into the same pass.  For
example, \citet{Sumii:2005aa} combines the first three of the above
passes into a single pass.

@@

The source language includes integers, Booleans, \texttt{Let} binding,
and conditional expressions. We choose a couple representative
primitive operators: \texttt{Sub} for integer subtraction and
\texttt{Eq} for equality on integers and Booleans.\footnote{Adding
more primitive operators is straightforward but uninteresting.}  The
\texttt{Read} operation requests an integer input from the user of the
program.\footnote{\texttt{Read} is necessary to prevent students from
interpreting the programs instead of compiling them.}

@@

The meaning of a whole program is given by the following relation.
\begin{lstlisting}
interp-x86-var : X86Var → Inputs → Value → Set
\end{lstlisting}

@@

\begin{lstlisting}
rco-correct-exp : ∀ (e : Exp) (ρ : Env Value)
  → interp-mon (rco e) ρ ≡ interp-exp e ρ
\end{lstlisting}

@@

\begin{lstlisting}
lift-locals : Mon → Imp-Prog
lift-locals m
    with lift-locals-mon m
... | n , e = Program n e    
\end{lstlisting}

@@

\begin{lstlisting}
$⇓$shifts : $∀${e}{v}{s $s′$}{$ρ₁$ $ρ′₁$ $ρ₂$ $ρ₃$ $ρ′₃$} 
  $→$ (s , $ρ₁$++$ρ₃$) $⊢$ e $⇓$ v $⊣$ ($s′$ , $ρ′₁$++$ρ′₃$)
  $→$ length $ρ′₁$ $≡$ length $ρ₁$
  $→$ (s , $ρ₁$++$ρ₂$++$ρ₃$) $⊢$
       shifts-imp-exp e (length $ρ₁$) (length $ρ₂$)
       $⇓$ v $⊣$ ($s′$ , $ρ′₁$++$ρ₂$++$ρ′₃$)
\end{lstlisting}

@@

All three functions will be building a control flow graph, which needs
to be threaded into and out of these functions, so they use the
following a \lstinline{Build} monad:

@@

The correctness theorem delegates to the correctness lemma for
\lstinline{select-stmt}, which says that if statement \lstinline{c}
evaluates to \lstinline{v}, then the instructions generated for
\lstinline{c} put the same value \lstinline{v} into \lstinline{rax}.
The list of x86 instruction blocks is obtain by mapping
\lstinline{select-stmt} over the control-flow graph \lstinline{G}.

\begin{lstlisting}
select-stmt-correct : $∀$ (c : CStmt) ($ρ$ $ρ′$ : Env Value)
  (s s$′$ : Inputs)(regs : List Value)(v : Value)(G : CFG)
  $→$ (s , $ρ$) , G $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
  $→$ 0 < length regs
  $→$ $∃$[ regs$′$ ] $∃$[ b ] (s , regs , $ρ$) , map select-stmt G
         $⊩$ select-stmt c $⇓$ (s$′$ , regs$′$ , $ρ′$) , b
    × nth regs$′$ rax $≡$ just v
\end{lstlisting}

@@

% LocalWords:  sigplan tex docstrip dtx bibtex TeXCount TC envir Siek
% LocalWords:  displaymath Nanopasses authornote authornotemark Tobin
% LocalWords:  rv Trovato al Agda nanopasses Olivier Danvy's ccs desc
% LocalWords:  nanopass Ichiro Aho qf gotos datatypes PEYTONJONES fk
% LocalWords:  Danvy arity Dybvig's Dybvig aa Chez Olivier's CompCert
% LocalWords:  fe Coq RTLgen CminorSel CompCert's CakeML stackLang tr
% LocalWords:  labLang Kumar ue Sarkar whac de Bruijn kx Eq Exp Num
% LocalWords:  Bool Var Int Env interp exp LIf immediates Atm atm mon
% LocalWords:  LMonIf IR AST Prog StateImp cmpq je jmp CExp CStmt CFG
% LocalWords:  IfEq CProg prog Arg Reg Dest Inst MovQ SubQ CmpQ JmpEq
% LocalWords:  ReadInt rax StateX EFLAGS var inst rco cmplx suc IL ie
% LocalWords:  retB thenB pred GHC thn eval Agda's transl condexpr
% LocalWords:  stmt arg dest regs DestOK trans

>>>>

Olivier Danvy, in his paper \emph{A New One-Pass Transformation into
Monadic Normal Form}~\citep{Danvy:2003fk}, presents an elegant family
of structurally recursive functions that translate Boolean expressions
and conditionals into a monadic normal form that (1) no longer
contains Boolean expressions, (2) the condition expression of a
condition is a value, and (3) jumps are expressed using zero-arity
function calls. One of the key parts of that translation is in the
compilation of a conditional expression such as $(\IF{b}{e_1}{e_2})$,
where the expression $b$ is compiled using a special function
$\mathcal{B}$ that takes two more arguments, the generated code for
the two branches, that is, for the two continuations of $b$. Thus,
$\mathcal{B}$ can analyze the expression $b$ to decide how to generate
code for the $\mathsf{if}$, or whether an $\mathsf{if}$ is even
necessary.
\[
\CE{\IF{b}{e_1}{e_2}} = \CB{b}{\CE{e_1}}{\CE{e_2}}
\]
For example, in the simple cases where $b$ is $\TRUE$ or $\FALSE$,
no branching is needed.
\begin{align*}
\CB{\TRUE}{c_1}{c_2} &= c_1 \\
\CB{\FALSE}{c_1}{c_2} &= c_2
\end{align*}
\noindent The treatment code as continuation goes back to earlier work
by \citet{Lawall:1994aa} on continuation-based partial
evaluation. Indeed, the above is a partial evaluation of Boolean
expressions.

My first exposure to these ideas was in Kent Dybvig's compiler
course~\citep{Dybvig:2010aa} where he taught the techniques used in
the Chez Scheme compiler~\citep{Dybvig:2006aa}.  One of those
techniques is called destination-driven code
generation~\citep{Dybvig:1990aa}, which adds an extra parameter to the
compilation function to indicate how the context will use the result
value of an expression. The context may (1) ignore the result (and
only care about the expression's side effects), (2) branch based on
interpreting its result as a boolean, (3) assign the result to a
variable, or (4) output the result. These options are encoded as a
disjoint union.  Relating this to Olivier's transformation functions,
the $\mathcal{E}$ function is for contexts that need the actual value
and the $\mathcal{B}$ function is for contexts that branch on the
expression's result. Of course, a pair of functions is isomorphic to a
function on a disjoint union.
\[
(A → C) × (B → C) ≅ (A + B) → C
\]

The CompCert compiler \citep{Leroy:2006fe} (a compiler for C verified
in Rocq) uses the one-function per context approach of Danvy in the
RTLgen pass that translates from the CminorSel intermediate language
to CompCert's register transfer language. The verified CakeML
compiler, on the other hand, employs a naive translation of Boolean
expressions and conditionals for the compilation of the stackLang 
intermediate language into labLang~\citep{Kumar:2014aa}.

@@

This paper presents a proof of correctness in Agda for four nanopasses
that translate a source language with let binding, integer arithmetic,
conditional expressions and Booleans into an x86-flavored register
transfer language. The Agda code is available in the supplementary
material. It does not depend on particular features of Agda and could
have been carried out in Rocq, Isabelle, or Lean.\footnote{The choice
of Agda for this paper is not an endorsement of Agda but is a
byproduct of the author's relative expertise in Agda at the time of
writing.}

The four compiler passes presented here correspond to the
following three passes in \emph{Essentials of Compilation}:
\begin{enumerate}
\item Remove Complex Operands
\item Explicate Control
\item Select Instructions
\end{enumerate}
The process of proving correctness of the Explicate Control pass
made it clear to me that it has two distinct responsibilities that
deserve to be separated:
\begin{itemize}
\item Replacing \texttt{let} expressions with assignment statements.
\item Translating Boolean expressions and conditionals into
  \texttt{goto}-based control-flow.
\end{itemize}
(The RTLgen pass of CompCert also handles the above two
responsibilities.)  So in this paper, the Explicate Control pass is
split into two passes, the first named Lift Locals and the second that
keeps the name Explicate Control.
\begin{enumerate}
\item Remove Complex Operands
\item Lift Locals
\item Explicate Control
\item Select Instructions
\end{enumerate}
In compilers that do not take the nanopass approach, unrelated
responsibilities are sometimes combined into the same pass.  For
example, \citet{Sumii:2005aa} combines the first three of the above
passes into a single pass.

@@

The source language includes integers, Booleans, \texttt{Let} binding,
and conditional expressions. We choose two representative primitive
operators: \texttt{Sub} for integer subtraction and \texttt{Eq} for
equality on integers and Booleans.\footnote{Adding more primitive
operators is straightforward but uninteresting.}  The \texttt{Read}
operation requests an integer input from the user of the
program.\footnote{\texttt{Read} is necessary to prevent students from
interpreting the programs instead of compiling them.}

@@

The meaning of a whole program is given by the following relation.
\begin{lstlisting}
interp-x86-var : X86Var $→$ Inputs $→$ Value $→$ Set
\end{lstlisting}

@@

\begin{lstlisting}
rco-correct-exp : $∀$ (e : Exp) ($ρ$ : Env Value)
  $→$ interp-mon (rco e) $ρ$ $≡$ interp-exp e $ρ$
\end{lstlisting}

@@

\begin{lstlisting}
lift-locals : Mon $→$ Imp-Prog
lift-locals m
    with lift-locals-mon m
... | n , e = Program n e    
\end{lstlisting}

@@

\begin{lstlisting}
$⇓$shifts : $∀${e}{v}{s $s′$}{$ρ₁$ $ρ′₁$ $ρ₂$ $ρ₃$ $ρ′₃$} 
  $→$ (s , $ρ₁$ ++ $ρ₃$) $⊢$ e $⇓$ v $⊣$ ($s′$ , $ρ′₁$ ++ $ρ′₃$)
  $→$ length $ρ′₁$ $≡$ length $ρ₁$
  $→$ (s , $ρ₁$ ++ $ρ₂$ ++ $ρ₃$)
      $⊢$ shifts-imp-exp e (length $ρ₁$) (length $ρ₂$)
      $⇓$ v $⊣$ ($s′$ , $ρ′₁$ ++ $ρ₂$ ++ $ρ′₃$)
\end{lstlisting}

@@

All three functions will be building a control flow graph, which needs
to be threaded into and out of these functions, so they use the
following \lstinline{Build} monad:

@@

The correctness theorem delegates to the correctness lemma for
\lstinline{select-stmt}, which says that if statement \lstinline{c}
evaluates to \lstinline{v}, then the instructions generated for
\lstinline{c} put the same value \lstinline{v} into \lstinline{rax}.
The list of x86 instruction blocks is obtained by mapping
\lstinline{select-stmt} over the control-flow graph \lstinline{G}.

\begin{lstlisting}[basicstyle=\ttfamily\footnotesize]
select-stmt-correct : $∀$ (c : CStmt) ($ρ$ $ρ′$ : Env Value)
  (s s$′$ : Inputs)(regs : List Value)(v : Value)(G : CFG)
  $→$ (s , $ρ$) , G $⊢ᶜ$ c $⇓$ v $⊣$ (s$′$ , $ρ′$)
  $→$ 0 < length regs
  $→$ $∃$[ regs$′$ ] $∃$[ b ] (s , regs , $ρ$) , map select-stmt G
         $⊩$ select-stmt c $⇓$ (s$′$ , regs$′$ , $ρ′$) , b
    × nth regs$′$ rax $≡$ just v
\end{lstlisting}

@@

% LocalWords:  sigplan tex docstrip dtx bibtex TeXCount TC envir Siek
% LocalWords:  displaymath Nanopasses authornote authornotemark Tobin
% LocalWords:  rv Trovato al Agda nanopasses Olivier Danvy's ccs desc
% LocalWords:  nanopass Ichiro Aho qf gotos datatypes PEYTONJONES fk
% LocalWords:  Danvy arity Dybvig's Dybvig aa Chez Olivier's CompCert
% LocalWords:  fe Rocq RTLgen CminorSel CompCert's CakeML stackLang tr
% LocalWords:  labLang Kumar ue Sarkar whac de Bruijn kx Eq Exp Num
% LocalWords:  Bool Var Int Env interp exp LIf immediates Atm atm mon
% LocalWords:  LMonIf IR AST Prog StateImp cmpq je jmp CExp CStmt CFG
% LocalWords:  IfEq CProg prog Arg Reg Dest Inst MovQ SubQ CmpQ JmpEq
% LocalWords:  ReadInt rax StateX EFLAGS var inst rco cmplx suc IL ie
% LocalWords:  retB thenB pred GHC thn eval Agda's transl condexpr
% LocalWords:  stmt arg dest regs DestOK trans
